An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-21 paper.tex
reorganize the paper

This paper is organized as follows. The evolving-graph BFS algorithm is described
and analyzed using graph notation in Section \ref{sec:breadth-first-search}. Section \ref{sec:evolv-graph-algor} considers the algebraic representation of the algorithm: 
A sparse matrix representation
of an evolving graph is presented in \ref{sec:representation}. 
The algorithm is presented in \ref{sec:breadth-first-search-1} and 
the properties of the search tree is discussed in \ref{sec:short-temp-path}.
The connection between a temporal path and a matrix-matrix multiplication is considered and discussed in \ref{sec:matr-matr-mult}. Finally, 
an application of the evolving-graph BFS in mining citation networks is described in Section \ref{sec:applications}.

@@

For the graph $G^{[i]} = (V^{[i]}, E^{[i]})$ at timestamp $i$, 
we denote the active node set of $V^{[i]}$ as $\tilde V^{[i]}$, 
where we have $\tilde V^{[i]} \subseteq V^{[i]}$.
Step $10$ of  Algorithm \ref{alg:bfs} ensures each active node-timestamp pair in $G_n$ is enqueued in $next$ (and then $frontier$) at most once. 
And so \texttt{forward-neighbors} loops through each 
active node-timestamp pair at most once. 
The operations of enqueuing take $O(1)$ time.
Thus the total operations of enqueuing are 
$O(\sum_{i=1}^n|\tilde V^{[i]}|)$, where $|\cdot|$ represents the number of elements of a set.  Let $\sum_i |\tilde V^{[i]}| = |\tilde V|$ and $\sum_i |E^{[i]}| = |\hat E|$.
Then the time of scanning
\texttt{forward-neighbors} is 
\[
\sum_{i=1}^n\sum_{v \in \tilde V^{[i]}} \texttt{forward-neighbors}(v) = 
O(|\hat E|),
\]
by the definition of forward-neighbors.
Hence the total running time is $O(|\hat E|+ |\tilde V|)$.

Note that the number of adjacency lists is the same as the number
of active nodes in $G_n$. So for large $n$, this representation can use a large amount of memory.
%Also need to consider the operations for matrix lists.

\section{Traversing Evolving Graphs Using Linear Algebraic Operations}
\label{sec:evolv-graph-algor}

\subsection{Representation}
\label{sec:representation}
As before, let $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$ be an evolving graph with $n$ timestamps. For each graph $G^{[t]}$, we use $A^{[t]}$ to denote
the corresponding adjacency matrix.
\[
A^{[t]}(i,j) = 
\begin{cases}
 1 & \mbox{if $(i,j)$ is an edge of $G^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]
 Each $A^{[t]}$ can be represented by 
a $|V|\times |V|$ sparse matrix,
where $|V|$ is the total number of nodes in $G_n$.
We use a sequence of adjacency matrices $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ to represent $G_n$. 
For example, the evolving graph in Figure \ref{fig:eg_shortest_path}
can be represented as 
\[
\Big\langle
  \begin{bmatrix} 
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0 
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0 
 \end{bmatrix}
\Big\rangle
\]
We assume $A_n$ is stored as an array of sparse matrices in compressed sparse column (CSC) format.

\subsection{Forward and Backward Searching}
\label{sec:breadth-first-search-1}

@@

@@

\subsection{Matrix-Matrix Multiplication}
\label{sec:matr-matr-mult}

@@

@@

In this paper, we have introduced and analyzed the evolving-graph BFS in both 
graph and algebraic notation. 
In order to study evolving graphs and experiment with various graph types, we have developed EvolvingGraphs.jl \cite{zhang15}, a software package in the Julia language \cite{bkse12} for the creation, manipulation, and study of dynamic networks.

This paper leaves important extensions open for further research. First, it is 
of interest to extend the evolving-graph BFS algorithm for weighted evolving graphs, which can provide more valuable information when the degree of connectivity is available in the data. 
Second, efficient algebraic representations and algorithms need to be addressed. 
One major advantage of representing evolving graphs as sparse matrices
is the matrix representation is highly structured so it is easy to design
high performance parallel implementations. 
Third, depth-first search and other graph traversal algorithms can be 
extended for searching evolving graphs using similar ideas.

>>>>

% This paper is organized as follows. The evolving-graph BFS algorithm is described
% and analyzed using graph notation in Section \ref{sec:breadth-first-search}. Section \ref{sec:evolv-graph-algor} considers the algebraic representation of the algorithm: 
% A sparse matrix representation
% of an evolving graph is presented in \ref{sec:representation}. 
% The algorithm is presented in \ref{sec:breadth-first-search-1} and 
% the properties of the search tree is discussed in \ref{sec:short-temp-path}.
% The connection between a temporal path and a matrix-matrix multiplication is considered and discussed in \ref{sec:matr-matr-mult}. Finally, 
% an application of the evolving-graph BFS in mining citation networks is described in Section \ref{sec:applications}.

@@

%Also need to consider the operations for matrix lists.

\subsection{Algebraic Evolving-Graph BFS}
\label{sec:representation}
As before, let $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$ be an evolving graph with $n$ timestamps. For each graph $G^{[t]}$, we use $A^{[t]}$ to denote
the corresponding adjacency matrix.
\[
A^{[t]}(i,j) = 
\begin{cases}
 1 & \mbox{if $(i,j)$ is an edge of $G^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]
 Each $A^{[t]}$ can be represented by 
a $|V|\times |V|$ sparse matrix,
where $|V|$ is the total number of nodes in $G_n$.
We use a sequence of adjacency matrices $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ to represent $G_n$. 
For example, the evolving graph in Figure \ref{fig:eg_shortest_path}
can be represented as 
\[
\Big\langle
  \begin{bmatrix} 
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0 
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0 
 \end{bmatrix}
\Big\rangle
\]

@@

% adjacency matrices
Suppose $G_n$ is represented by adjacency lists.
Then the number of adjacency lists is the same as the number
of active nodes in $G_n$. So for large $n$, this representation 
can use a large amount of memory.
For the graph $G^{[i]} = (V^{[i]}, E^{[i]})$ at timestamp $i$, 
we denote the active node set of $V^{[i]}$ as $\tilde V^{[i]}$, 
where we have $\tilde V^{[i]} \subseteq V^{[i]}$.
Step $10$ of  Algorithm \ref{alg:bfs} ensures each active node-timestamp pair in $G_n$ is enqueued in $next$ (and then $frontier$) at most once. 
And so \texttt{forward-neighbors} loops through each 
active node-timestamp pair at most once. 
The operations of enqueuing take $O(1)$ time.
Thus the total operations of enqueuing are 
$O(\sum_{i=1}^n|\tilde V^{[i]}|)$, where $|\cdot|$ represents the number of elements of a set.  Let $\sum_i |\tilde V^{[i]}| = |\tilde V|$ and $\sum_i |E^{[i]}| = |\hat E|$.
Then the time of scanning
\texttt{forward-neighbors} is 
\[
\sum_{i=1}^n\sum_{v \in \tilde V^{[i]}} \texttt{forward-neighbors}(v) = 
O(|\hat E|),
\]
by the definition of forward-neighbors.
Hence the total running time is $O(|\hat E|+ |\tilde V|)$.

@@

\section{Node Activeness}
\label{sec:matr-matr-mult}

\subsection{Temporal Path}
\label{sec:temporal-path}

In this subsection, we compare with the ``wrong'' way of traveling an evolving graph
and show the importance of node activeness.

\subsection{Matrix-Matrix Multiplication}
\label{sec:matr-matr-mult-1}

@@

\section{Implementation in Julia}
\label{sec:implementation-julia}

In order to study evolving graphs and experiment with various graph types, we have developed EvolvingGraphs.jl \cite{zhang15}, a software package in the Julia language \cite{bkse12} for the creation, manipulation, and study of dynamic networks.

In Julia,  $A_n$ is stored as an array of sparse matrices in compressed sparse column (CSC) format.

@@

In this paper, we have introduced and analyzed the evolving-graph BFS in both 
graph and algebraic notation. 
This paper leaves important extensions open for further research. First, it is 
of interest to extend the evolving-graph BFS algorithm for weighted evolving graphs, which can provide more valuable information when the degree of connectivity is available in the data. 
Second, efficient algebraic representations and algorithms need to be addressed. 
One major advantage of representing evolving graphs as sparse matrices
is the matrix representation is highly structured so it is easy to design
high performance parallel implementations. 
Third, depth-first search and other graph traversal algorithms can be 
extended for searching evolving graphs using similar ideas.
